---
title: "Data Wrangling - Manipulation"
output: html_document
---

```{r}

library(tidyverse)

options(tibble.print_min = 3)

litters_data = read_csv("./data/data_import_examples/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/data_import_examples/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

##Select variables (picking what columns you want)

```{r}
select(litters_data, group, litter_number, gd0_weight)

select(litters_data, group, litter_number, gd0_weight, pups_survive)


```

...by specifying a range

```{r}
select(litters_data, gd_of_birth:pups_survive)

#can skip around
select(litters_data, group, gd_of_birth:pups_survive)

```

...by saying what you don't want to keep

```{r}
select(litters_data, -litter_number)
```

...and changing the variable names

```{r}
select(litters_data, group, litter_number, gest_day_0_weight = gd0_weight)

#if you just want to rename a column in the original dataset
rename(litters_data, gest_day_0_weight = gd0_weight)
```

...use select helpers

```{r}
select(litters_data, starts_with('gd')) #column names that start with gd
select(litters_data, group, starts_with('pup')) #mix group variable with variables that start with gd

select(litters_data, litter_number, everything()) #reorganize your variables
```

Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears

```{r}
select(pups_data, litter_number, sex, pd_ears)
```


##Filter observations (picking what rows you want)

```{r}

```





Pipes are easier to understand and avoids clutter (avoid nested functions/operations)
