---
title: "Data Wrangling - Manipulation"
output: html_document
---

```{r}

library(tidyverse)

options(tibble.print_min = 3)

litters_data = read_csv("./data/data_import_examples/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/data_import_examples/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

##Select variables (picking what columns you want)

```{r}
select(litters_data, group, litter_number, gd0_weight)

select(litters_data, group, litter_number, gd0_weight, pups_survive)


```

...by specifying a range

```{r}
select(litters_data, gd_of_birth:pups_survive)

#can skip around
select(litters_data, group, gd_of_birth:pups_survive)

```

...by saying what you don't want to keep

```{r}
select(litters_data, -litter_number)
```

...and changing the variable names

```{r}
select(litters_data, group, litter_number, gest_day_0_weight = gd0_weight)

#if you just want to rename a column in the original dataset
rename(litters_data, gest_day_0_weight = gd0_weight)
```

...use select helpers (rename)

```{r}
select(litters_data, starts_with('gd')) #column names that start with gd
select(litters_data, group, starts_with('pup')) #mix group variable with variables that start with gd

select(litters_data, litter_number, everything()) #reorganize your variables
```

Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears

```{r}
select(pups_data, litter_number, sex, pd_ears)
```


##Filter observations (picking what rows you want)

Do some filtering
```{r}
filter(litters_data, gd0_weight < 25)

filter(litters_data, group %in% c("Con7", "Con8"))
filter(litters_data, group == "Con7" | group == "Con8")

filter(litters_data, pups_born_alive == 8)

filter(litters_data, !is.na(gd0_weight))

filter(litters_data, pups_born_alive %in% 4:6, !is.na(gd0_weight))
```

Learning Assessment: Filter to include only pups with sex 1 and filter to include only pups with PD walk less than 11 and sex 2
```{r}
filter(pups_data, sex == 1)
filter(pups_data, pd_walk < 11, sex == 2)
```


##Mutate

Create variables

```{r}
mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       group = tolower(group) #change variable name to lower case
       )

mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       wt_gain_squared = wt_gain^2)


```


##Arrange

Arrange the data

```{r}
arrange(litters_data, gd0_weight)

arrange(litters_data, desc(pups_born_alive), gd0_weight)

```


##Using pipes

Example of intermediate object approach (not easy to read):

```{r}
litters_data_raw = read_csv("./data/data_import_examples/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = mutate(litters_data_selected_cols, 
  wt_gain = gd18_weight - gd0_weight,
  group = tolower(group))
```

Example of nested object approach (not easy to read):

```{r}
  mutate(
    select(
      janitor::clean_names(
        read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii")
        ), 
    -pups_survive
    ),
  wt_gain = gd18_weight - gd0_weight,
  group = tolower(group)
  )
```

Use piping:

```{r}
litters_data = 
  read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = tolower(group))

```

Pipes are easier to understand and avoids clutter (avoid nested or intermediate functions/operations). %>% takes the output in the current line and becomes the first input/argument in the next line. 

If output is not the first argument in the next function (use a "." as placeholder) :

```{r}
litters_data = 
  read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = tolower(group)
    ) %>%
  filter(!is.na(gd0_weight)) %>%
  lm(gd18_weight ~ gd0_weight, data = .) #use . to specify where the argument should go
 
```

There are limitations to piping - they are extremely sequential: would not work if trying to clean two datasets and merge them.