---
title: "Tidy Data"
author: "Margaret Gacheru"
date: "September 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Import pulse data

```{r}
library(tidyverse)
pulse_data = haven::read_sas("./data/data_import_examples/public_pulse_data.sas7bdat") %>%
  janitor::clean_names()
pulse_data
```

With our new understanding of tidy data, we quickly recognize a problem: the BDI score is spread across four columns, which correspond to four observation times. We want each patient to have four rows (turn column names into observations). 

```{r}
pulse_tidy_data = gather(pulse_data, key = visit, value = bdi, bdi_score_bl:bdi_score_12m)
```

Illustrate 'separate - split visit into separate columns
```{r}
separate(pulse_tidy_data, visit, into = c("bdi_str", "score_str", "visit"), sep = "_") %>%
  select(-bdi_str, -score_str) %>%
  mutate(visit = replace(visit, visit == "bl", "00m"))

```

Putting it all together - doing all the data cleaning at once

```{r}
pulse_data = haven::read_sas("./data/data_import_examples/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  gather(key = visit, value = bdi, bdi_score_bl:bdi_score_12m) %>%
  separate(visit, into = c("remove_1", "remove_2", "visit"), sep = "_") %>%
  select(id, visit, everything(), -starts_with("remove")) %>%
  mutate(visit = replace(visit, visit == "bl", "00m"))
```


Import the litters data; separate group into dose and day after location 3

```{r}
read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>%
  mutate(dose = tolower(dose),
         wt_gain = gd18_weight - gd0_weight) %>%
  arrange(litter_number)

```

In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only  litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 

```{r}
read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(litter_number, ends_with("weight")) %>%
  gather(key = "gest_day", value = "weight", gd0_weight:gd18_weight) %>%
  mutate(gest_day = recode(gest_day, "gd0_weight" = 0, "gd18_weight" = 18)) %>%
  arrange(litter_number)

```

Opposite of gather - untidy your tidy data (making this human readable)

```{r}
analysis_result = tibble(
  group = c("treatment", "treatment", "placebo", "placebo"),
  time = c("pre", "post", "pre", "post"),
  mean = c(4, 8, 3.5, 4)
) %>%
  spread(key = "time", value = "mean") %>% #value is the observation and key is the column 
  knitr::kable()

```

Binding rows

```{r}
fellowship_ring = readxl::read_excel("./data/data_import_examples/LotR_Words.xlsx", range = "B3:D6") %>%
  mutate(movie = "fellowship_ring")

two_towers = readxl::read_excel("./data/data_import_examples/LotR_Words.xlsx", range = "F3:H6") %>%
  mutate(movie = "two_towers")

return_king = readxl::read_excel("./data/data_import_examples/LotR_Words.xlsx", range = "J3:L6") %>%
  mutate(movie = "return_king")
```

Put them together in a way that keeps everything
```{r}
lotr_tidy = bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names() %>%
  gather(key = "sex", value = "words", female:male) %>%
  mutate(race = tolower(race)) %>% 
  select(movie, everything()) 
```


Join data sets
```{r}
pup_data = read_csv("./data/data_import_examples/FAS_pups.csv", col_types = "ciiiii") %>%
  janitor::clean_names() %>%
  mutate(sex = recode(sex, `1` = "male", `2` = "female")) 

litter_data = read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = tolower(group))

FAS_data = left_join(pup_data, litter_data, by = "litter_number")
FAS_data
```

